A weakly structured stem for human origins in Africa

### 2022-07-15 paper.tex
Addressing Reviewer #3 point #3a

\begin{table}[t!]
    \centering
    \scriptsize
    \begin{tabular}{lllrrrr}
        \toprule
        \textbf{Likelihood} & \textbf{Label} & \textbf{Population Pair} &
            \textbf{Divergence} & \textbf{Migration rate} & \textbf{Migration} \\
        & & & 
            \textbf{Time (ka)} & \textbf{per generation} & \textbf{duration (kyr)} \\
        \midrule
        \textbf{Continuous Model} & & & & & \\
        $LL= -115,500$ & a & Stem 1, Stem 2 & 1,163 & 6.43e-5 & 1,028 \\
        & b & Stem 2, Nama & NA & 5.82e-5 & 130 \\
        & c, d & Stem 2, Other Africans* & NA & 3.10e-5, \textbf{1.64e-4} & 130, 55 \\
        & e, f & Nama, Other Africans* & 135 & 4.1e-5, 9.8e-6 & 135, 60 \\
        & g & Mende, East Africans & 60 & \textbf{2.14e-4} & 60 \\
        & h & East Africans, British & 50 & 4.17e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo & 12 &\textbf{3.36e-4} & 12 \\
        \textbf{Merger Model} & & & & & \\
        $LL= -102,600$ & a & Stem 1, Stem 2 & 1,442 & \textbf{1.16e-4} & 963 \\
        & -- & Stem 1S, Stem 1E & 479 & 0 (fixed) & -- \\
        & b & Stem 2 to Nama & 119 & \textbf{0.71} & pulse \\
        & c & Stem 2 to Stem 1E & 98 & \textbf{0.50} & pulse \\
        & d & Stem 2 to Mende & 25 & \textbf{0.18} & pulse \\
        & e, f & Nama, Other Africans* & 119 & 4.4e-5, 7.1e-6 & 119, 60 \\
        & g & Mende, East Africans* & 60 & \textbf{1.98e-4} & 60 \\
        & h & East Africans, British & 50 & 3.87e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo &12 & \textbf{3.59e-4} & 12 \\
        \bottomrule
    \end{tabular}
    \caption{
        \textbf{Migration and divergence parameters from best fit models.}
        Labeled migration rates correspond to symmetric continuous migration
        bands shown in Figure~\ref{fig:best-fit-models}. Both the continuous migration and
        merger models inferred a relatively deep split of human stem branches,
        though they were connected by ongoing migration that maintained their
        genetic similarity. Bold text indicates migration rates above $10^{-4}$.
        In both models, the branch ancestral to the Nama shares a common ancestral population
        with the other African groups $\sim$120--135ka. Following this divergence,
        the population ancestral to other African groups branches into West and East African
        groups 60ka.
        *Migration rates and durations are shown between branches ancestral to 
        1) Nama and East Africans and their ancestors, and
        2) Nama and Mende, respectively.
        ``Divergence times'' correspond to the most recent common ancestral population,
        and does not account for continuous migration or earlier reticulations.
    }
    \label{tab:migration-rates}
\end{table}

@@

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups. 
We observe a sharp bottleneck in Stem 1 down to $N_e=117$ after the split of
the Neanderthal branch. This represents the lower bound allowed in our
optimization (i.e. 1\% of the ancestral $N_e$), although the size of this
bottleneck is poorly constrained ($\sigma_{N_e}=838$). After a long period of
exchange with Stem 2, Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S''
479ka. The timing of this divergence was also poorly constrained ($\sigma_T=
166$ka). These populations evolve independently until approximately 119ka
when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
equal proportions (50\% each) to form the ancestors of the western Africans and
eastern Africans (and thus also all individuals who later disperse during the
Out of Africa event). Finally, the Mende receive a large additional pulse of
gene flow from Stem 2, replacing 18\% of their population 25ka.
The later Stem 2 contribution to the western African Mende resulted in
better model fits ($\Delta LL \approx 60,000$).
Speculatively, this may indicate that an ancestral Stem 2 population occupied western or
Central Africa, broadly speaking. The differing proportions in the Nama and
eastern Africans may also indicate geographic separation of Stem 1S in southern
Africa and Stem 1E in eastern Africa.

@@

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S (Figure~\ref{fig:best-fit-models}B) 
from southern Africa moved northward thereby encountering Stem 2,
or Stem 2 moved from central/western Africa southward into Stem 1S -- then we
could observe disproportionate ancestry contributions from different stems in
contemporary groups. We observed two merger events. The first, between Stem 1S and
Stem 2, results in the formation of an ancestral Khoe-San population ~120ka.
The second $\approx100$ka between Stem 1E and Stem 2, results in the formation of the
ancestors of East/West Africans as well as later ``Out of Africans''. The rapid
rise in sea levels and increased precipitation during MIS 5e, following a
glacial period of aridity across Africa \citep{Blome2012-lw}, might have
triggered migration inland away from the coasts, as has been suggested, e.g., for the
Paleo-Agulhas plain \citep{Marean2014-pg}.

>>>>

\begin{table}[t!]
    \centering
    \scriptsize
    \begin{tabular}{lllrrrr}
        \toprule
        \textbf{Likelihood} & \textbf{Label} & \textbf{Population Pair} &
            \textbf{Divergence} & \textbf{Migration rate} & \textbf{Migration} \\
        & & & 
            \textbf{Time (ka)} & \textbf{per generation} & \textbf{duration (kyr)} \\
        \midrule
        \textbf{Continuous Model} & & & & & \\
        $LL= -115,500$ & a & Stem 1, Stem 2 & 1,163 & 6.43e-5 & 1,028 \\
        (Table S4) & b & Stem 2, Nama & NA & 5.82e-5 & 130 \\
        & c, d & Stem 2, Other Africans* & NA & 3.10e-5, \textbf{1.64e-4} & 130, 55 \\
        & e, f & Nama, Other Africans* & 135 & 4.1e-5, 9.8e-6 & 135, 60 \\
        & g & Mende, East Africans & 60 & \textbf{2.14e-4} & 60 \\
        & h & East Africans, British & 50 & 4.17e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo & 12 &\textbf{3.36e-4} & 12 \\
        \textbf{Merger Model} & & & & & \\
        $LL= -102,600$ & a & Stem 1, Stem 2 & 1,442 & \textbf{1.16e-4} & 963 \\
        (Table S6) & -- & Stem 1S, Stem 1E & 479 & 0 (fixed) & -- \\
        & b & Stem 2 to Nama & 119 & \textbf{0.71} & pulse \\
        & c & Stem 2 to Stem 1E & 98 & \textbf{0.50} & pulse \\
        & d & Stem 2 to Mende & 25 & \textbf{0.18} & pulse \\
        & e, f & Nama, Other Africans* & 119 & 4.4e-5, 7.1e-6 & 119, 60 \\
        & g & Mende, East Africans* & 60 & \textbf{1.98e-4} & 60 \\
        & h & East Africans, British & 50 & 3.87e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo &12 & \textbf{3.59e-4} & 12 \\
        \bottomrule
    \end{tabular}
    \caption{
        \textbf{Migration and divergence parameters from best fit models.}
        Labeled migration rates correspond to symmetric continuous migration
        bands shown in Figure~\ref{fig:best-fit-models}. Both the continuous migration and
        merger models inferred a relatively deep split of human stem branches,
        though they were connected by ongoing migration that maintained their
        genetic similarity. Bold text indicates migration rates above $10^{-4}$.
        In both models, the branch ancestral to the Nama shares a common ancestral population
        with the other African groups $\sim$120--135ka. Following this divergence,
        the population ancestral to other African groups branches into West and East African
        groups 60ka.
        *Migration rates and durations are shown between branches ancestral to 
        1) Nama and East Africans and their ancestors, and
        2) Nama and Mende, respectively.
        ``Divergence times'' correspond to the most recent common ancestral population,
        and does not account for continuous migration or earlier reticulations.
    }
    \label{tab:migration-rates}
\end{table}

@@

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups. 
We observe a sharp bottleneck in Stem 1 down to $N_e=117$ after the split of
the Neanderthal branch. This represents the lower bound allowed in our
optimization (i.e. 1\% of the ancestral $N_e$), although the size of this
bottleneck is poorly constrained ($\sigma_{N_e}=838$). After a long period of
exchange with Stem 2, Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S''
479ka. The timing of this divergence was also poorly constrained ($\sigma_T=
166$ka). These populations evolve independently until 119ka
($\sigma_T=
28$ka) when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
equal proportions (50\% each) to form the ancestors of the western Africans and
eastern Africans (and thus also all individuals who later disperse during the
Out of Africa event). Finally, the Mende receive a large additional pulse of
gene flow from Stem 2, replacing 18\% of their population 25ka ($\sigma_T=
0.6$ka).
The later Stem 2 contribution to the western African Mende resulted in
better model fits ($\Delta LL \approx 60,000$).
Speculatively, this may indicate that an ancestral Stem 2 population occupied western or
Central Africa, broadly speaking. The differing proportions in the Nama and
eastern Africans may also indicate geographic separation of Stem 1S in southern
Africa and Stem 1E in eastern Africa.

@@

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S (Figure~\ref{fig:best-fit-models}B) 
from southern Africa moved northward thereby encountering Stem 2,
or Stem 2 moved from central/western Africa southward into Stem 1S -- then we
could observe disproportionate ancestry contributions from different stems in
contemporary groups. We observed two merger events. The first, between Stem 1S and
Stem 2, results in the formation of an ancestral Khoe-San population $\approx120$ka.
The second $\approx100$ka between Stem 1E and Stem 2, results in the formation of the
ancestors of East/West Africans as well as later ``Out of Africans''. The rapid
rise in sea levels and increased precipitation during MIS 5e, following a
glacial period of aridity across Africa \citep{Blome2012-lw}, might have
triggered migration inland away from the coasts, as has been suggested, e.g., for the
Paleo-Agulhas plain \citep{Marean2014-pg}.
